---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# SamSeq

Working with aligned reads from a `*.sam` file.

Allows loading `*.sam` files, i.e. the format used by SamTools. Tracks header information, reads, and the source of the file. Loads all reads and optionally the extra read tags into S3 objects that are also data frames. This allows easy manipulation and rquires much less overhead than using Bioconductor, but is significantly slower and uses more memory.

Includes functions to translate the Sam flags and to filter reads.

### Loading a sam file

A `Sam` object is created/constructed by the `Sam()` function, usually by passing it the name of a `*.sam` file as a parameter. This object is a list with a `SamHeader` and a `SamRead` object, and has a `SamSource` object as an attribute. These can be accessed directly but should use the accessor/creator methods of the same name as the object classes. The folowing code uses a tiny sam file distributed with this package.

```{r}
library(SamSeq)

# Get the name of a *. sam file to use in this example. If SamSeq is not
# installed in the normal place, specify the R library directory with
# the system.file parameter lib.loc=
sample.sam <- system.file( "extData/pe.sam", package = "SamSeq" )

samObj <- Sam( sample.sam )
class(samObj)
names(samObj)
names(attributes(samObj))
```

The SamHeader and SamReads objects are also data frames, the SamSource object
is also a list.

```{r}
samHeaderObj <- SamHeader( samObj )
class(samHeaderObj)
samReadsObj <- SamReads( samObj )
class(samReadsObj)
samSourceObj <- SamSource( samObj )
class(samSourceObj)
```

### SamHeader objects

The SamHeader is not fully parsed at this point, it will probably have a different structure in future. For now it is provided as a data frame with only two columns: the initial tag is parsed, as the column `tag`, and the rest of the line is provided uparsed, as the column `record`.

The problem yet to be determined is how to split the contents when some header lines use tabs for delimiters and others can contain embedded unquoted tabs. Probably some form of parameter object that will allow specifying, on a per tag basis, what to do.

```{r}
samHeaderObj <- SamHeader( samObj )
names(samHeaderObj)
samHeaderObj[1, ]
samHeaderObj$tag
samHeaderObj[samHeaderObj$tag == "SQ", "record" ]
```

